Computable Compressed Matrices\\Suplementary Material

### 2014-04-04 suppl.tex
revisei o suplemento e o Manuscrito. Deixei alguns todos no Suplemento para informação. por favor remova @crysttian

In this supplementary material we will explore basic arithmetic operations
(addition, subtraction, division and multiplication) over the elements of
bitstring compressed arrays. We will consider integer matrices compressed via
both the SM and the VLB methods, \todo{inseri esse texto} but the proposed 
methods can be applied to integers and real numbers with sign.

@@

It also presents an application of the developed library. The results of 
operations implemented and implementation are presented in a matrix problem of 
collaborative filtering.\todo{Inseri para colocar a parte da implementação.}

@@

\begin{table}[ht]
	\centering
    \caption{Comparing binary and decimal sum ($7+10$).}
    \begin{tabular}{crrrc}
    \hline
    	  & Decimal & & Binary & bit-length \\
    \hline      
    Carry Over& 	& & \textit{11100} & \\
    $n_1$ & 7   & & 111 & 3 \\
    $n_2$ & \underline{10}	& & \underline{1010} & 4 \\
    Total & 17	& & 10001 & 5 \\
    \hline
	\end{tabular}
    \label{tab:01}
\end{table}
\todo{Traduzi o título que estava em português.}
\begin{table}[ht]
    \centering
    \caption{Comparison of the sum in binary and decimal base.}
    \begin{tabular}{crrrc}
    \hline
    	  & Decimal & & Binario & Bit-length \\
    \hline      
    Carry & 	& & \textit{1110} & \\
    $n_1$ & 7   & & 111 & 3 \\
    $n_1$ & \underline{~7}	& & \underline{111} & 3 \\
    Total & 14	& & 1110 & 4 \\
    \hline
	\end{tabular}
    \label{tab:02}
\end{table}

@@

Depending of the number involved in this operation, the result may be negative.
This fact alone generates the need for an extra bit to represent the sign ($+$
or $-$). The usual way to achieve this is to use the  two's complement
\cite{flores1963logic} representation which has the advantage of making the
addition subtraction and multiplication operation the same as those for unsigned
binary numbers. This algorithm \todo{removi is what} is actually implemented in 
CPUs. To exemplify let's subtract $7$ from $10$. To make use of the
same algorithm of the sum , we represent the operation as $10+(-7)$. First we
need to convert the operands to two's complement representation (C2):

@@

\todo{adicionei esse trecho - As can be seen in} As can be seen in figure 
\ref{fig:bitlength}, the bit-lengths of results of the four
operations with integer operands up to 8 bits in length. The conclusions listed
above, are visually emphasized in the figures.

@@

To calculate $P[1,1]$ which in uncompressed form is $1 \times 1 + 3 \times 3 + 5
\times 7 + 8 \times 9$, we first must extract the numbers from the bitstrings
and then proceed with the linear combination. The extraction will make use of
masks as described before. Let $tempA$ and $tempB$ store the values of blocks 1 
and 3 of AA[1,1] and BB[1,1]. Now we need to determine the size of the blocks
(bit-length) containing each element of the resulting matrix. As demonstrated
before, in the worst case, the product will have a bit-length which is the sum
of the bit-lengths of the operands. For this example this length is $4+5=9$,
however, we also have three additions for each element,adding a total of 3
extra bits, thus we end up with a bit-length of 12 for each element of the
product. Having determined the bit-length of the product, we can now \todo{we 
can - removi} do the actual calculations and store the results.

@@

This paper proposes the bitstring-compression methodologies. The SM 
method was implemented as a Fortran library which is available under an 
open-source license\todo{colocar o link para o repositório aqui}\todo{Acho 
melhor mandar para o artigo e anexar a biblioteca, para depois disponibilizar 
o código.}. The 
implementation extends the standard Matrix type of Fortran, overloading 
operators such as assignment, addition, subtraction, multiplication, transpose, 
and maximum of a matrix. 
Although only few significant matrix operations have been ovearloaded in this 
example implementation, any other operation can be performed, since 
the methods for inserting and collecting an element in a matrix of bitstring 
were also implemented. With these two methods, as long as they are properly 
adapted into algorithms, it is possible to implement any other desired 
operation.

The SM method was also extended to allow for the allocation of real and 
signed integer matrices, besides positive integer ones. Some adjustments were 
made so that the other types of numbers could be represented. In order to 
represent the signed integers, it was necessary to use two's complement, as it 
is applied to thein accordance to traditional methods\todo{acho que esta 
explicação ficou muito solta. Não precisamos explica a implementação no 
artigo, podemos colocar na documentação do código}. \todo{Eu deixaria, pois é 
apenas um apontamento com relação as operações, pois no arquivo que 
compartilhei com você eu coloco os códigos fontes com comentários.}

@@

The results are represented in Figures \ref{fig39}, \ref{fig40} and \ref{fig41} 
for the PUA Method and in Figures 
\ref{fig42}, \ref{fig43} and \ref{fig44} for the 
BFM Methods.  A computer with 8 GB of RAM Memory, 16 GB of swap, processor core 
i7 model 870  \todo{confirmar modelo}
and operational 
system Linux Debian 7.0 was used to perform this simulation. In Figure 
\ref{fig39}, it is possible to verify the approximate 
moment  when the  the storage in disk (swaap) begins to be used. This is 
indicated by the vertical green line. 
At this moment, the bitstring method becomes faster than the traditional method 
because it does not require disk 
storage due tocompressed memory representation. Note that, in the Figure 
\ref{fig40}, the traditional method is faster 
than the bitstring one before the beginning of the disk access. However, after 
it becomes necessary use the disk 
so that the matrix can be accessed and stored, , the traditional method  becames 
slower (Figura \ref{fig41}). 
These results 
indicate that the bistring method can promote the optimization of the memory use 
because the access to the main 
memories (RAM and cache) are faster than the access to secondary memories 
(disk). In Figures \ref{fig42}, 
\ref{fig43} and \ref{fig44} the 
results of the processing time for  the BFM method are represented. Once again, 
a vertical green line indicates 
the approximate moment when the disk access begins. There is a difference in 
relation to the PUA method since the 
BMF method uses the PUA values to perform its calculation. Therefore, the BMF 
method demands more memory to be 
processed, which reduces the matrix dimensions involved in the process, 
represented in the main memory. As 
indicated in Figure \ref{fig41}, for the PUA method, and Figure \ref{fig44}, for 
the BMF method, it is possible 
to verify that 
before the access to the disk, the traditional method is faster than the 
bitstring one, but once the access 
is initiated,  the bitstring method becames faster.

>>>>

In this supplementary material we will explore basic arithmetic operations
(addition, subtraction, division and multiplication) over the elements of
bitstring compressed arrays. We will consider integer matrices compressed via
both the SM and the VLB methods, but the proposed 
methods can be applied to integers and real numbers with sign.

@@

It also presents an application of the developed library. We apply our 
implementation to classical machine learning problem: collaborative filtering.

@@

\begin{table}[ht]
	\centering
    \caption{Comparing binary and decimal sum ($7+10$).}
    \begin{tabular}{crrrc}
    \hline
    	  & Decimal & & Binary & bit-length \\
    \hline      
    Carry Over& 	& & \textit{11100} & \\
    $n_1$ & 7   & & 111 & 3 \\
    $n_2$ & \underline{10}	& & \underline{1010} & 4 \\
    Total & 17	& & 10001 & 5 \\
    \hline
	\end{tabular}
    \label{tab:01}
\end{table}
\todo{Traduzi o título que estava em português.}
\begin{table}[ht]
    \centering
    \caption{Comparing binary and decimal sum of 7+7.}
    \begin{tabular}{crrrc}
    \hline
    	  & Decimal & & Binary & Bit-length \\
    \hline      
    Carry & 	& & \textit{1110} & \\
    $n_1$ & 7   & & 111 & 3 \\
    $n_1$ & \underline{~7}	& & \underline{111} & 3 \\
    Total & 14	& & 1110 & 4 \\
    \hline
	\end{tabular}
    \label{tab:02}
\end{table}

@@

Depending of the number involved in this operation, the result may be negative.
This fact alone generates the need for an extra bit to represent the sign ($+$
or $-$). The usual way to achieve this is to use the  two's complement
\cite{flores1963logic} representation which has the advantage of making the
addition subtraction and multiplication operation the same as those for unsigned
binary numbers. This algorithm is actually implemented in 
CPUs. To exemplify let's subtract $7$ from $10$. To make use of the
same algorithm of the sum , we represent the operation as $10+(-7)$. First we
need to convert the operands to two's complement representation (C2):

@@

As can be seen in figure 
\ref{fig:bitlength}, the bit-lengths of results of the four
operations with integer operands up to 8 bits in length. The conclusions listed
above, are visually emphasized in the figures.

@@

To calculate $P[1,1]$ which in uncompressed form is $1 \times 1 + 3 \times 3 + 5
\times 7 + 8 \times 9$, we first must extract the numbers from the bitstrings
and then proceed with the linear combination. The extraction will make use of
masks as described before. Let $tempA$ and $tempB$ store the values of blocks 1 
and 3 of AA[1,1] and BB[1,1]. Now we need to determine the size of the blocks
(bit-length) containing each element of the resulting matrix. As demonstrated
before, in the worst case, the product will have a bit-length which is the sum
of the bit-lengths of the operands. For this example this length is $4+5=9$,
however, we also have three additions for each element,adding a total of 3
extra bits, thus we end up with a bit-length of 12 for each element of the
product. Having determined the bit-length of the product, we can now do the 
actual calculations and store the results.

@@

This paper proposes the bitstring-compression methodologies. The SM 
method was implemented as a Fortran library which is available under an 
open-source license. The 
implementation extends the standard Matrix type of Fortran, overloading 
operators such as assignment, addition, subtraction, multiplication, transpose, 
and maximum of a matrix. 
Although only few significant matrix operations have been ovearloaded in this 
example implementation, any other operation can be performed, since 
the methods for inserting and collecting an element in a matrix of bitstring 
were also implemented. With these two methods, as long as they are properly 
adapted into algorithms, it is possible to implement any other desired 
operation.

The SM method was also extended to allow for the allocation of real and 
signed integer matrices, besides positive integer ones. Some adjustments were 
made so that the other types of numbers could be represented. \todo{removi 
``in order to represent...''}

@@

The results are represented in Figures \ref{fig39}, \ref{fig40} and \ref{fig41} 
for the PUA Method and in Figures 
\ref{fig42}, \ref{fig43} and \ref{fig44} for the 
BFM Methods.  A computer with 8 GB of RAM Memory, 16 GB of swap, processor core 
i7 model 870
and operational 
system Linux Debian 7.0 was used to perform this simulation. In Figure 
\ref{fig39}, it is possible to verify the approximate 
moment  when the  the storage in disk (swaap) begins to be used. This is 
indicated by the vertical green line. 
At this moment, the bitstring method becomes faster than the traditional method 
because it does not require disk 
storage due tocompressed memory representation. Note that, in the Figure 
\ref{fig40}, the traditional method is faster 
than the bitstring one before the beginning of the disk access. However, after 
it becomes necessary use the disk 
so that the matrix can be accessed and stored, , the traditional method  becames 
slower (Figura \ref{fig41}). 
These results 
indicate that the bistring method can promote the optimization of the memory use 
because the access to the main 
memories (RAM and cache) are faster than the access to secondary memories 
(disk). In Figures \ref{fig42}, 
\ref{fig43} and \ref{fig44} the 
results of the processing time for  the BFM method are represented. Once again, 
a vertical green line indicates 
the approximate moment when the disk access begins. There is a difference in 
relation to the PUA method since the 
BMF method uses the PUA values to perform its calculation. Therefore, the BMF 
method demands more memory to be 
processed, which reduces the matrix dimensions involved in the process, 
represented in the main memory. As 
indicated in Figure \ref{fig41}, for the PUA method, and Figure \ref{fig44}, for 
the BMF method, it is possible 
to verify that 
before the access to the disk, the traditional method is faster than the 
bitstring one, but once the access 
is initiated,  the bitstring method becames faster.
